---
title: "Sentiment"
author: "Winston Saunders"
date: "May 9, 2016"
output: 
    html_document:
        theme: united
---
<style>
  .col2 {
    columns: 2 300px;         /* number of columns and width in pixels */
    -webkit-columns: 2 300px; /* chrome, safari */
    -moz-columns: 2 300px;    /* firefox */
  }
  .col3 {
    columns: 3 200px;
    -webkit-columns: 3 200px;
    -moz-columns: 3 200px;
  }
</style>

<style>
tr:hover {background-color: #BBFFFF}
table { 
    width: 80%;
    display: table;
    border-collapse: collapse;
    border-spacing: 18px;
    border-color: #AAAAFF;
    background-color: #AFEEEE;
    padding: 2px;
    font: 12px arial, sans-serif;
}
th, td{
    text-align: center;
}
</style>

###SUMMARY
 
This continuation of [BIAS AND CONTEXT IN PRESIDENTIAL DEBATE TEXTS](https://rpubs.com/ww44ss/Debate_Text) and [HEAT MAPS & LINGUISTICS OF DEBATE SPEECH in R](rpubs.com/ww44ss/HeatMapRDebates), which focused on a "Bag of Words" approach to analyzing the text of Presidential Debates for the 2016 Election Cycle.

This extends the analysis to include "sentiment" analysis based on standard semiment models using the {virdis} package in R.

###DATA SOURCES AND METHODS
The text of the presidential debates are downloaded from the [UCSB Presidency Project](http://www.presidency.ucsb.edu/debates.php). Transcripts were pasted into Apple Pages and stored as unformatted .txt files.  




```{r, "find debate files in directory", echo=FALSE, warning=FALSE, message=FALSE}

    ## 
    ## .txt file detection
    

    directory <- "/Users/winstonsaunders/Documents/presidential_debates_clinton_saunders/"
    file.list <- list.files(directory)
    
    republican.files <- file.list[grepl("Republican", file.list)]
    democratic.files <- file.list[grepl("Democratic", file.list)]
    
    
    republican.file.names <- republican.files
    democratic.file.names <- democratic.files
    
    
    republican.files <- paste0(directory, republican.files)
    democratic.files <- paste0(directory, democratic.files)

```

```{r, "load local functions", echo=FALSE}

## LOAD LOCAL FUNCTIONS

    ## load text of debate files
    source(paste0(directory, "load_debate_text.R"))
    ## select candidate text
    source(paste0(directory, "candidate_text.R"))
    ## turn candidate text in a Text Corpus
    source(paste0(directory, "text_tc.R"))
    ## selects candidate text and then turns that into a Text Corpus
    source(paste0(directory, "candidate_text_tc.R"))
    ## enable printing of multiple plots
    source(paste0(directory, "multiplot.R"))
    ## vector normialize
    source(paste0(directory, "vector.normalize.R"))

```

```{r, "load r libraries", echo=FALSE, warning=FALSE, message=FALSE}

    ## check for installed packages
    if(! ("tm" %in% rownames(installed.packages())))      {install.packages("tm")}
    if(! ("RWeka" %in% rownames(installed.packages())))   {install.packages("RWeka")}
    if(! ("SnowballC" %in% rownames(installed.packages()))) {install.packages("SnowballC")}
    if(! ("ggplot2" %in% rownames(installed.packages()))) {install.packages("ggplot2")}
    if(! ("xtable" %in% rownames(installed.packages())))  {install.packages("xtable")}
    if(! ("reshape2" %in% rownames(installed.packages())))  {install.packages("reshape2")}
    if(! ("dplyr" %in% rownames(installed.packages())))  {install.packages("dplyr")}

    library(dplyr)
    library(tm)
    library(RWeka)
    library(SnowballC)
    library(ggplot2)
    library(xtable)
    library(reshape2)
    library(stringr)
    if(! ("stringr" %in% rownames(installed.packages())))  {install.packages("stringr")}
    if(! ("syuzhet" %in% rownames(installed.packages())))  {install.packages("syuzhet")}
    if(! ("viridis" %in% rownames(installed.packages())))  {install.packages("viridis")}
    library(stringr)
    library(viridis)
    library(syuzhet)

    

```

```{r, "get text files", echo=FALSE, warning=FALSE, message=FALSE}


    ## THIS CODE CHUNK GETS ALL THE CANDIDATE TEXT FROM FILES STORED LOCALLY

    ## create dummy files for all republicans
    r_all <- NULL
    r.names.list <- c("QUINTANILLA", "KASICH", "HUCKABEE", "BUSH", "RUBIO", "TRUMP", "CARSON", "FIORINA", "CRUZ", "CHRISTIE", "PAUL", "HARWOOD", "QUICK", "CRAMER", "SANTELLI", "EPPERSON",                    "TAPPER", "DINAN","BASH","HEWITT", "CAVUTO", "BARTIROMO","BAKER","UNKNOWN","BLITZER","JOSH JACOB, COLLEGE STUDENT", "QUESTION" ,"AUDIENCE", "BAIER", "KELLY", "WALLACE","MUIR", "RADDATZ", "HAM" , "MCELVEEN", "DICKERSON", "STRASSEL", "GARRETT", "ARRARÃS", "CUOMO", "HANNITY", "UNIDENTIFIABLE"  )


    for (file_name in republican.files){ 
        ## load the text
        temp <- file_name %>% load_debate_text
        
        ## extract date info from file name
        date.regex <- "[A-z]+ [0-9]{1,2} [0-9]{4}"
        date.text <- str_extract(file_name, date.regex)
        
        temp_tibl_df <- data_frame("debate" = date.text, "name" = temp$name, text = temp$text)
        
        
        ## clean up and add index
            temp_tibl_df <- temp_tibl_df[temp_tibl_df$name %in% r.names.list,]
            temp_tibl_df$speaking_index <- seq.int(nrow(temp_tibl_df))
        
        r_all <-  r_all %>% rbind(temp_tibl_df)
    }
    
    
    #d.names.list <- c( "HOLT","CLINTON","SANDERS","MITCHELL", "FRANTA", "BROWNLEE", "DICKERSON", "CORDES", "COONEY", "WEBB", "BASH", "LEMON", "WILKINS", "LOPEZ", "RADDATZ", "LEVESQUE", "MCELVEEN" )
    
    
    d.names.list <- c("COOPER", "CHAFEE","WEBB","OMALLEY","SANDERS", "CLINTON" ,"BASH", "LEMON", "WILKINS", "LOPEZ", "OMALLEY ", "QUESTION", "DICKERSON", "CORDES", "COONEY", "OBRADOVICH", "RADDATZ", "MUIR", "ANNOUNCER", "LEVESQUE", "MCELVEEN", "HOLT", "FRANCHESCA RAMSEY", "MITCHELL","FRANTA", "BROWNLEE", "TODD", "MADDOW", "WOODRUFF", "IFILL", "RAMOS", "SALINAS ", "RAMOS ", "SALINAS", "TUMULTY", "QUESTION ", "SMITH", "UNKNOWN ", "UNKNOWN", "BLITZER", "LOUIS", "AUDIENCE MEMBER") 
    
    
    d_all<-NULL
    
    for (file_name in democratic.files){ 
        ## load the text
        temp <- file_name %>% load_debate_text
        
        ## extract date info from file name
        date.regex <- "[A-z]+ [0-9]{1,2} [0-9]{4}"
        date.text <- str_extract(file_name, date.regex)
        
        temp_tibl_df <- data_frame("debate" = date.text, "name" = temp$name, text = temp$text)
        
        
        ## clean up and add index
            temp_tibl_df <- temp_tibl_df[temp_tibl_df$name %in% d.names.list,]
            temp_tibl_df$speaking_index <- seq.int(nrow(temp_tibl_df))
        
        d_all <-  d_all %>% rbind(temp_tibl_df)
    }
    
    
    
    d_all_f <- as_data_frame(d_all)
    d_all_f$concat_index <- seq.int(nrow(d_all_f))
    
    r_all_f <- as_data_frame(r_all)
    r_all_f$concat_index <- seq.int(nrow(r_all_f))

```

```{r, "dummify commentators", echo=FALSE}


d.candidates.list <- c("CHAFEE", "WEBB", "CLINTON", "SANDERS", "OMALLEY")

r.candidates.list <- c("KASICH", "HUCKABEE", "BUSH", "RUBIO", "CARSON", "FIORINA", "CRUZ", "CHRISTIE", "PAUL", "TRUMP", "FIORINA")


## keep just candidate names
r_all_f <- r_all_f %>% mutate(candidate.name = ifelse(name %in% r.candidates.list, name, "OTHER") )
## keep just candidate names
d_all_f <- d_all_f %>% mutate(candidate.name = ifelse(name %in% d.candidates.list, name, "OTHER") )



```

####Democratic Debates

```{r, ech0=FALSE}

print(unique(d_all_f$debate))

```


#### Republican Debates

```{r, ech0=FALSE}

print(unique(r_all_f$debate))


```

###Compute Sentiment


```{r}


## add sentiment
d_all_f <- mutate(d_all_f, sentiment = get_sentiment(d_all_f$text, method = "afinn"))

r_all_f <- mutate(r_all_f, sentiment = get_sentiment(r_all_f$text, method = "afinn"))


```


```{r, fig.align="center", fig.width=9, fig.height=5, echo=FALSE}

p <- ggplot(d_all_f, aes(x = concat_index, y = sentiment, fill=debate, color=debate)) + 
    geom_bar(stat='identity') + 
    facet_grid(candidate.name~.)
    #geom_smooth(span = 0.3, n = 100)

p 
```

```{r, fig.align="center", fig.width=9, fig.height=7, echo=FALSE}


p <- ggplot(r_all_f, aes(x = concat_index, y = sentiment, fill=debate, color=debate)) + 
    geom_bar(stat="identity") + 
    facet_grid(candidate.name~.) 
p 



```





```{r, fig.width=5, fig.height=8, echo=FALSE, fig.align='center'}

trump_all <- r_all_f %>% filter(candidate.name=="TRUMP")

p <- ggplot(trump_all, aes(x = speaking_index, y = sentiment)) + 
    geom_bar(stat="identity", color = ) + 
    facet_grid(debate~.) 
p 


```

```{r, fig.width=5, fig.height=8, echo=FALSE, fig.align='center'}

p <- ggplot(r_all_f[r_all_f$candidate.name == "CRUZ",], aes(x = speaking_index, y = sentiment)) + 
    geom_bar(stat="identity", color = "salmon") + 
    facet_grid(debate~.) 
p 


```