---
title: "Sentiment"
author: "Winston Saunders"
date: "May 9, 2016"
output: 
    html_document:
        theme: united
---
<style>
  .col2 {
    columns: 2 200px;         /* number of columns and width in pixels */
    -webkit-columns: 2 200px; /* chrome, safari */
    -moz-columns: 2 200px;    /* firefox */
  }
  .col3 {
    columns: 3 200px;
    -webkit-columns: 3 200px;
    -moz-columns: 3 200px;
  }
</style>

<style>
tr:hover {background-color: #FFEE75}
table { 
    width: 40%;
    display: table;
    border-spacing: 18px;
    border-color: #AAAAFF;
    background-color: #FFFFFF;
    padding: 2px;
    font: 14px arial, sans-serif;
}
th, td{
    text-align: center;
    border: 1px solid black;
    border-collapse: collapse;
}
</style>

###SUMMARY
 
This continuation of [BIAS AND CONTEXT IN PRESIDENTIAL DEBATE TEXTS](https://rpubs.com/ww44ss/Debate_Text) and [HEAT MAPS & LINGUISTICS OF DEBATE SPEECH in R](rpubs.com/ww44ss/HeatMapRDebates), which focused on a "Bag of Words" approach to analyzing the text of Presidential Debates for the 2016 Election Cycle.

This extends the analysis to include "sentiment" analysis based on standard semiment models using the {virdis} package in R.

##DATA SOURCES AND METHODS
The text of the presidential debates are downloaded from the [UCSB Presidency Project](http://www.presidency.ucsb.edu/debates.php). Transcripts were pasted into Apple Pages and stored as unformatted .txt files.  

```{r, "find debate files in directory", echo=FALSE, warning=FALSE, message=FALSE}

    ## 
    ## .txt file detection
    

    directory <- "/Users/winstonsaunders/Documents/presidential_debates_clinton_saunders/"
    file.list <- list.files(directory)
    
    republican.files <- file.list[grepl("Republican", file.list)]
    democratic.files <- file.list[grepl("Democratic", file.list)]
    
    
    republican.file.names <- republican.files
    democratic.file.names <- democratic.files
    
    
    republican.files <- paste0(directory, republican.files)
    democratic.files <- paste0(directory, democratic.files)

```

```{r, "load local functions", echo=FALSE}

## LOAD LOCAL FUNCTIONS

    ## load text of debate files
    source(paste0(directory, "load_debate_text.R"))
    ## select candidate text
    source(paste0(directory, "candidate_text.R"))
    ## turn candidate text in a Text Corpus
    source(paste0(directory, "text_tc.R"))
    ## selects candidate text and then turns that into a Text Corpus
    source(paste0(directory, "candidate_text_tc.R"))
    ## enable printing of multiple plots
    source(paste0(directory, "multiplot.R"))
    ## vector normialize
    source(paste0(directory, "vector.normalize.R"))

```

```{r, "load r libraries", echo=FALSE, warning=FALSE, message=FALSE}

    ## check for installed packages

    


    currently.installed.packages <- rownames(installed.packages()) 


    #package.library <- function(pkg = "ggplot2") {
    #    if(! (pkg %in% currently.installed.packages))   {install.packages(pkg, character.only = TRUE)}
    #    library(pkg)
    #}
    
    

    if(! ("tm" %in% currently.installed.packages))        {install.packages("tm")}
    if(! ("RWeka" %in% currently.installed.packages))     {install.packages("RWeka")}
    if(! ("SnowballC" %in% currently.installed.packages)) {install.packages("SnowballC")}
    if(! ("ggplot2" %in% currently.installed.packages))   {install.packages("ggplot2")}
    if(! ("xtable" %in% currently.installed.packages))    {install.packages("xtable")}
    if(! ("reshape2" %in% currently.installed.packages))  {install.packages("reshape2")}
    if(! ("dplyr" %in% currently.installed.packages))     {install.packages("dplyr")}
    if(! ("stringr" %in% currently.installed.packages))   {install.packages("stringr")}
    if(! ("syuzhet" %in% currently.installed.packages))   {install.packages("syuzhet")}
    if(! ("viridis" %in% currently.installed.packages))   {install.packages("viridis")}

    #required.packages <- c("tm", "dplyr", "RWeka", "SnowballC", "ggplot2", "reshape2", "stringr", "viridis", "syuzhet")
    
    #lapply(required.packages, package.library)
    
    library(dplyr)
    library(tm)
    library(RWeka)
    library(SnowballC)
    library(ggplot2)
    library(xtable)
    library(reshape2)
    library(stringr)
    library(viridis)
    library(syuzhet)
    library(xtable)

    

```

```{r, "get text files", echo=FALSE, warning=FALSE, message=FALSE}


    ## THIS CODE CHUNK GETS ALL THE CANDIDATE TEXT FROM FILES STORED LOCALLY

    ## create dummy files for all republicans
    r_all <- NULL
    r.names.list <- c("QUINTANILLA", "KASICH", "HUCKABEE", "BUSH", "RUBIO", "TRUMP", "CARSON", "FIORINA", "CRUZ", "CHRISTIE", "PAUL", "HARWOOD", "QUICK", "CRAMER", "SANTELLI", "EPPERSON", "TAPPER", "DINAN","BASH","HEWITT", "CAVUTO", "BARTIROMO","BAKER","UNKNOWN","BLITZER","JOSH JACOB, COLLEGE STUDENT", "QUESTION" ,"AUDIENCE", "BAIER", "KELLY", "WALLACE","MUIR", "RADDATZ", "HAM" , "MCELVEEN", "DICKERSON", "STRASSEL", "GARRETT", "ARRARÃS", "CUOMO", "HANNITY", "UNIDENTIFIABLE"  )


    for (file_name in republican.files){ 
        ## load the text
        temp <- file_name %>% load_debate_text
        
        ## extract date info from file name
        date.regex <- "[A-z]+ [0-9]{1,2} [0-9]{4}"
        date.text <- str_extract(file_name, date.regex)
        
        temp_tibl_df <- data_frame("debate" = date.text, "name" = temp$name, text = temp$text)
        
        
        ## clean up and add index
            temp_tibl_df <- temp_tibl_df[temp_tibl_df$name %in% r.names.list,]
            temp_tibl_df$speaking_index <- seq.int(nrow(temp_tibl_df))
        
        r_all <-  r_all %>% rbind(temp_tibl_df)
    }
    
    
    #d.names.list <- c( "HOLT","CLINTON","SANDERS","MITCHELL", "FRANTA", "BROWNLEE", "DICKERSON", "CORDES", "COONEY", "WEBB", "BASH", "LEMON", "WILKINS", "LOPEZ", "RADDATZ", "LEVESQUE", "MCELVEEN" )
    
    
    d.names.list <- c("COOPER", "CHAFEE","WEBB","OMALLEY","SANDERS", "CLINTON" ,"BASH", "LEMON", "WILKINS", "LOPEZ", "OMALLEY ", "QUESTION", "DICKERSON", "CORDES", "COONEY", "OBRADOVICH", "RADDATZ", "MUIR", "ANNOUNCER", "LEVESQUE", "MCELVEEN", "HOLT", "FRANCHESCA RAMSEY", "MITCHELL","FRANTA", "BROWNLEE", "TODD", "MADDOW", "WOODRUFF", "IFILL", "RAMOS", "SALINAS ", "RAMOS ", "SALINAS", "TUMULTY", "QUESTION ", "SMITH", "UNKNOWN ", "UNKNOWN", "BLITZER", "LOUIS", "AUDIENCE MEMBER") 
    
    
    d_all<-NULL
    
    for (file_name in democratic.files){ 
        ## load the text
        temp <- file_name %>% load_debate_text
        
        ## extract date info from file name
        date.regex <- "[A-z]+ [0-9]{1,2} [0-9]{4}"
        date.text <- str_extract(file_name, date.regex)
        
        temp_tibl_df <- data_frame("debate" = date.text, "name" = temp$name, text = temp$text)
        
        
        ## clean up and add index
            temp_tibl_df <- temp_tibl_df[temp_tibl_df$name %in% d.names.list,]
            temp_tibl_df$speaking_index <- seq.int(nrow(temp_tibl_df))
        
        d_all <-  d_all %>% rbind(temp_tibl_df)
    }
    
    
    
    d_all_f <- as_data_frame(d_all)
    d_all_f$concat_index <- seq.int(nrow(d_all_f))
    
     ## reorder factor levels of debates
    
    d_all_f$debate <- as.factor(d_all_f$debate)
    d_all_f$debate <- factor(d_all_f$debate,levels(d_all_f$debate)[c(9, 8, 2, 5, 4, 3, 6, 7,1)])
    
    r_all_f <- as_data_frame(r_all)
    r_all_f$concat_index <- seq.int(nrow(r_all_f))
    
    ## reorder factor levels of debates
    
    r_all_f$debate <- as.factor(r_all_f$debate)
    r_all_f$debate <- factor(r_all_f$debate,levels(r_all_f$debate)[c(10, 9, 1, 5, 6, 4, 2, 3, 8, 7)])

```

```{r, "dummify commentators", echo=FALSE}

d.candidates.list <- c("CHAFEE", "WEBB", "CLINTON", "SANDERS", "OMALLEY")

r.candidates.list <- c("KASICH", "HUCKABEE", "BUSH", "RUBIO", "CARSON", "FIORINA", "CRUZ", "CHRISTIE", "PAUL", "TRUMP", "FIORINA")


## keep just candidate names
r_all_f <- r_all_f %>% mutate(candidate.name = ifelse(name %in% r.candidates.list, name, "OTHER") )
## keep just candidate names
d_all_f <- d_all_f %>% mutate(candidate.name = ifelse(name %in% d.candidates.list, name, "OTHER") )



```

##DEMOCRAITC DEBATES

Here are the dates of the analyzed Democratic Debates

<style>
table { 
    width: 200px;
    margin-left: auto;
    margin-right: auto;
    }
</style>

```{r, ech0=FALSE, results='asis'}

print(xtable(as.matrix(unique(d_all_f$debate)), align = c("c", "c")), type="html", include.rownames=FALSE, include.colnames = FALSE)



```

###Republican Debates

Here are the dates of the analyzed Republican Debates

```{r, ech0=FALSE, results='asis'}

print(xtable(as.matrix(unique(r_all_f$debate)), align = c("c", "c")), type="html", include.rownames=FALSE, include.colnames = FALSE)


```

##Use `mutate()` to add computed sentiment  

We'll use the `mutate()` to add a "sentiment" column to the data. In this case I'll just use two of the standard methods "bing" and "afinn" included in the package. Each of these associates a specific "sentiment" to a set of words, then simply sums the result. 

```{r}
## add sentiment using two different methods (afinn and bing)
d_all_f <- mutate(d_all_f, afinn_sentiment = get_sentiment(d_all_f$text, method = "afinn"), bing_sentiment = get_sentiment(d_all_f$text, method = "bing"))

r_all_f <- mutate(r_all_f, afinn_sentiment = get_sentiment(r_all_f$text, method = "afinn"), bing_sentiment = get_sentiment(r_all_f$text, method = "bing"))

```


### Democratic Debate Sentiment

The plot below shows the trend of the overall sentiment of the all debates and all candidates. The "other" category lumps all audience and moderator comments together.  

```{r, fig.align="center", fig.width=9, fig.height=5, echo=FALSE, warning=FALSE}

    p <- ggplot(d_all_f, aes(x = concat_index, y = bing_sentiment, fill=debate, color=debate)) + 
    geom_bar(stat='identity') + 
    facet_grid(candidate.name~.)

    print(p)
```


Notice that the candidate sentiments have high variablity than do the moderators (as might be expected), with the debate of February 11 2016 standing out. [Press Coverage](http://www.cnn.com/2016/02/11/politics/democratic-debate-highlights/) of the February 11, 2016 debate revealed the following comments:   
_Much of the debate lacked the bitterness...[but] the confrontation during the PBS "NewsHour" Democratic debate simulcast on CNN flared into open anger in the final moments._
_Clinton accused her rival of not standing with Obama after he endorsed a book by CNN contributor Bill Press critical of the president. She said Sanders had called Obama "weak" and a "disappointment" in the past and she warned "the kind of criticism that we heard from Sen. Sanders about our president, I expect from Republicans. I do not expect (it) from someone running for the Democratic nomination to succeed President Obama."_  
  
Let's see if we can pick this up....  

###Use `filter()` to drill into February 11 2016  
Again we can use `{dplyr}` functions to get the data and then use the transform function in the `{syuzhet}` package to extract low frequency components.


```{r, fig.align="center", fig.height=4, fig.width=6, warning=FALSE}
candidate_x <- "CLINTON"
debate_x <- "February 11 2016"
single_d_debate <- d_all_f %>% filter(debate == debate_x, name == candidate_x)
filtered.sentiment = get_dct_transform(single_d_debate$afinn_sentiment, x_reverse_len=nrow(single_d_debate), low_pass_size = 7) 

tt <- filtered.sentiment %>% as.data.frame()
ttx <- cbind(tt, single_d_debate$afinn_sentiment)
colnames(ttx) <- c("filtered", "sentiment")

p <- ggplot(ttx, aes(y = filtered, x=seq_along(filtered))) +
    geom_bar(aes(y=sentiment), stat='identity', fill="salmon", color = "grey60", alpha=0.8) +
    geom_line(color = "blue", size = 1.5) + 
    ggtitle(paste(candidate_x, "on", debate_x))+
    xlab("sequence") +
    ylab("sentiment")

p

```

<style>
table { 
    width: 800px;
    margin-left: auto;
    margin-right: auto;
    }
</style>

```{r, echo=FALSE, results='asis'}

## create temp table of interesting data
temptable<-xtable(cbind(c(39:41),single_d_debate$text[39:41], single_d_debate$afinn_sentiment[39:41], single_d_debate$bing_sentiment[39:41]))
colnames(temptable) <- c("seq", "text", "afinn", "bing")
print(temptable, type="html", size="small", include.rownames=FALSE)

```
  
Indeed, we see a sharp negative trend in Hillary Clinton's bing sentiment score at the end of the debate which parallels the observation in the press above. 



## Republican Debate Sentiment

There were are a lot more candidates in the Republican filed, so the density of data is lower, especially in the earlier debates, where there was a plethora of them.

```{r, fig.align="center", fig.width=9, fig.height=7, echo=FALSE, warning=FALSE}


p <- ggplot(r_all_f, aes(x = concat_index, y = bing_sentiment, fill= debate, color=debate)) + 
    geom_bar(stat = 'identity') + 
    facet_grid(candidate.name~.) 
p 



```


###Donald Trump's Penis March 3, 2016   
The March 3, 2016 Republican debate will forever live in infamy because it struck possibly the lowest note of a presidential debate, ever. The press headlines read [Donald Trump defends size of his penis](http://www.cnn.com/2016/03/03/politics/donald-trump-small-hands-marco-rubio/) in coverage of a remarkable exchange with Marco Rubio. 
  
Again we can use `filter()` to get the data and then use the transform function in the `{syuzhet}` package to extract low frequency components.


```{r, fig.align="center", fig.height=4, fig.width=6, warning=FALSE}
candidate_x <- "TRUMP"
debate_x <- "March 3 2016"
single_r_debate <- r_all_f %>% filter(debate == debate_x, name == candidate_x)
filtered.sentiment = get_dct_transform(single_r_debate$afinn_sentiment, x_reverse_len=nrow(single_r_debate), low_pass_size = 10) 

tt <- filtered.sentiment %>% as.data.frame()
ttx <- cbind(tt, single_r_debate$afinn_sentiment)
colnames(ttx) <- c("filtered", "sentiment")

p <- ggplot(ttx, aes(y = filtered, x=seq_along(filtered))) +
    geom_bar(aes(y=sentiment/max(sentiment)), stat='identity', fill="salmon", color = "grey60", alpha=0.8) +
    geom_line(color = "blue", size = 1.5) + 
    ggtitle(paste(candidate_x, "on", debate_x))+
    xlab("sequence") +
    ylab("sentiment")

p

```

<style>
table { 
    width: 800px;
    margin-left: auto;
    margin-right: auto;
    }
</style>

```{r, echo=FALSE, results='asis'}

## create temp table of interesting data
seq_table <- 14:16

temptable<-xtable(cbind(c(seq_table),single_r_debate$text[seq_table], single_r_debate$afinn_sentiment[seq_table], single_r_debate$bing_sentiment[seq_table]))
colnames(temptable) <- c("seq", "text", "afinn", "bing")
print(temptable, type="html", size="small", include.rownames=FALSE)

```
  

###Using `filter()` to plot a histogram of specific candidate sentiment

####SANDERS

```{r, echo=TRUE}
sanders_sent <- d_all_f %>% filter(name == "SANDERS")
```

overall the distribution is pretty uninteresting

```{r, fig.align='center', fig.width=6, fig.height=4}
ggplot(sanders_sent, (aes(x = afinn_sentiment))) + geom_histogram(binwidth=2,fill="#467254")
```

however, if we look at the densities of individual debates we can see some subtle differences. 

```{r, fig.align='center', fig.width=6, fig.height=4}
ggplot(sanders_sent, aes(x=afinn_sentiment, ..density.., color = debate)) +
  geom_freqpoly(binwidth = 2, size=1.5)
```


In this case we can see differences between the debates. Particularly the Feb 4, 2016 debate. We can dig a little deeper by looking at the text

```{r}

sanders_high <- filter(d_all_f, name== "SANDERS", debate == "February 11 2016", afinn_sentiment >= 5, afinn_sentiment <= 7)
sanders_high$text[c(1, 3)]

```


####CRUZ

```{r, echo=TRUE}
can_sent <- r_all_f %>% filter(name == "CRUZ")
```


```{r}
ggplot(can_sent, aes(x=afinn_sentiment, ..density.., color = debate)) +
  geom_freqpoly(binwidth = 2, size=1.5)
```


```{r}
cruz_low <- filter(r_all_f, name== "CRUZ", debate == "December 15 2015", afinn_sentiment >= -6, afinn_sentiment <= -4)

cruz_low$text[5:6]
```



##Use `group_by()` to look at summary debate stats

####Democrats

We can look at the mean sentiment expressed in each debate by the two leading Democratic candidates.

```{r}
dem_stat <- group_by(d_all_f, name, debate)
summarize(dem_stat, mean_afinn = mean(afinn_sentiment), mean_bing = mean(bing_sentiment)) %>% filter(name == "CLINTON")

summarize(dem_stat, mean_afinn = mean(afinn_sentiment), mean_bing = mean(bing_sentiment)) %>% filter(name == "SANDERS")
```

Here is a plot of the data

```{r}
clinton <- summarize(dem_stat, mean_afinn = mean(afinn_sentiment), mean_bing = mean(bing_sentiment)) %>% filter(name == "CLINTON")

sanders <- summarize(dem_stat, mean_afinn = mean(afinn_sentiment), mean_bing = mean(bing_sentiment)) %>% filter(name == "SANDERS")

ggplot(rbind(clinton, sanders), aes(x = debate, y = mean_afinn, fill = name)) + geom_point(pch = 23, size = 4, color = "blue") + geom_point(size = 3, pch = 21, aes(y= 2*mean_bing))+theme(axis.text.x = element_text(angle=60, hjust=1)) + ggtitle("2*bing and afinn debate sentiment")


```

These correlate pretty well. 

```{r}

ggplot(d_all_f %>% filter(name == "CLINTON"), aes(x = afinn_sentiment, y = 2*bing_sentiment)) + geom_jitter(pch=21, width=2, height=2, alpha=.4, fill = "darkred") + geom_smooth(formula = y ~ x) + ggtitle("afinn versus 2*bing sentiment for Hillary Clinton")

```


####Republican

```{r, echo=FALSE}
rep_stat <- group_by(r_all_f, name, debate)

a <- summarize(rep_stat, mean_afinn = mean(afinn_sentiment), mean_bing = mean(bing_sentiment)) %>% filter(name == "TRUMP")

b <- summarize(rep_stat, mean_afinn = mean(afinn_sentiment), mean_bing = mean(bing_sentiment)) %>% filter(name == "CRUZ")

c <- summarize(rep_stat, mean_afinn = mean(afinn_sentiment), mean_bing = mean(bing_sentiment)) %>% filter(name == "RUBIO")

ggplot(rbind(a,b), aes(x = debate, y = mean_afinn, fill = name)) + geom_point(pch=22, size = 5, color = "blue") + theme(axis.text.x = element_text(angle=60, hjust=1))



```

### Polarized Statements

What statements were the most polarized (i.e. most positive and negative) of each candidate. We can slice this using filter and summarize.

####Hillary Clinton
#####Max Positive Sentiment (afinn)

```{r, echo=2}

hillary_max <- d_all_f %>% filter(name == "CLINTON") %>% filter(afinn_sentiment == max(afinn_sentiment))

hillary_max_text <- hillary_max$text

print(hillary_max_text)

```

#####Max Positive Sentiment (bing)

```{r, echo=2}

hillary_max <- d_all_f %>% filter(name == "CLINTON") %>% filter(afinn_sentiment == max(bing_sentiment))

hillary_max_text <- hillary_max$text

print(hillary_max_text)

```
#####Max Negative Sentiment (afinn)
```{r, echo=FALSE}

hillary_min <- d_all_f %>% filter(name == "CLINTON") %>% filter(afinn_sentiment == min(afinn_sentiment))

hillary_min_text <- hillary_min$text

print(hillary_min_text)

```

####Donald Trump
#####Max Positive Sentiment (afinn)

```{r, echo=2}

donald_max <- r_all_f %>% filter(name == "TRUMP") %>% filter(afinn_sentiment == max(afinn_sentiment))

print(donald_max$text)

```

#####Max Positive Sentiment (bing)

```{r, echo=2}

donald_max <- r_all_f %>% filter(name == "TRUMP") %>% filter(afinn_sentiment == max(bing_sentiment))

print(donald_max$text)

```
#####Max Negative Sentiment (afinn)
```{r, echo=FALSE}

donald_min <- r_all_f %>% filter(name == "TRUMP") %>% filter(afinn_sentiment == min(bing_sentiment))

print(donald_min$text)

```


```{r, fig.width=5, fig.height=8, echo=FALSE, fig.align='center'}

trump_all <- r_all_f %>% filter(candidate.name=="TRUMP")

p <- ggplot(trump_all, aes(x = speaking_index, y = afinn_sentiment)) + 
    geom_bar(stat="identity", color = "blue") + 
    facet_grid(debate~.) 
p 


```

```{r, fig.width=5, fig.height=8, echo=FALSE, fig.align='center'}

p <- ggplot(r_all_f[r_all_f$candidate.name == "CRUZ",], aes(x = speaking_index, y = afinn_sentiment)) + 
    geom_bar(stat="identity", color = "salmon") + 
    facet_grid(debate~.) 
p 


```